Supplementary Information: Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests (Anderson-Teixeira et al., *GCB*)

### 2021-10-25 doc/manuscript/ForestGEO_treerings.Rmd
one more orcid ID
add a couple ORCID IDs
ORCID
knit in figures with updated SCBI allometries
fix a few numbers following scbi allometry update

Tala Awada^8^

@@

Stuart J. Davies^2^

Cameron Dow^1,14^

@@

Sean McMahon^2,19^  (Orcid ID : 0000-0001-8302-6908)

@@

Anastasia E. Sniderhan^7^

@@

7. Biology Department, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5

@@

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to environmental drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter at breast height (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees usually more sensitive. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached maxima at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

### Most influential climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb11)
n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
n_DBHx_analyzed_sig=90  #NIO water is included here
#precip:
n_DBHxP_RW_analyzed= 36
n_DBHxP_RW_sig=16
n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 11)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb11)
n_tested=37 #same for all metrics

n_best_model_RW=31
n_best_model_BAI=32
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
percent_neg_of_sig_trends="90-91%"
percent_sig_neg_trend_of_analyzed = "73-78%" 
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

>>>>

Tala Awada^8^ ( Orcid ID : 0000-0002-0462-2339)

@@

Stuart J. Davies^2^ ( Orcid ID : 0000-0002-8596-7522)

Cameron Dow^1,14^ ( Orcid ID : 0000-0002-8365-598X)

@@

Sean M. McMahon^2,19^  (Orcid ID : 0000-0001-8302-6908)

@@

Anastasia E. Sniderhan^7^ ( Orcid ID : 0000-0003-4823-3180)

@@

7. Department of Biology, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5

@@

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to environmental drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter at breast height (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
Climate sensitivity commonly varied with DBH (45% of cases tested), with larger trees usually more sensitive. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached maxima at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

### Most influential climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Oct 25,2021)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Oct 25,2021)
n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
n_DBHx_analyzed_sig=91  #NIO water is included here
#precip:
n_DBHxP_RW_analyzed= 36
n_DBHxP_RW_sig=16
n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Oct 25,2021)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43
n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Oct 25,2021)
n_tested=37 #same for all metrics

n_best_model_RW=31
n_best_model_BAI=32
n_best_model_AGB=31
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
percent_neg_of_sig_trends="90-91%"
percent_sig_neg_trend_of_analyzed = "73-78%" 
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
